Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2024-10-31 ms.qmd
Updates

## Abstract
Behavioural flexibility is expected to be crucial for animals adapting to environmental changes. However, the conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how populations will respond to novel circumstances. Stressful situations faced by the parents can have a direct impact on offspring cognition through the transmission of glucocorticoids - stress-related hormones that affect offspring cognition. At the same time, stressful conditions can influence parental behaviour during nesting and consequently the thermal developmental conditions offspring experience. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. We hypothesized prenatal CORT and cold temperatures would impair performance in the reversal task. Given _L. delicata_'s success as an invasive species, we expected this species to behave more flexibly and be less affected by early environmental conditions. Contrary to our expectations, behavioural flexibility appears to be robust to prenatal temperature and CORT in both species. The lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that novel environments are unlikely to influence flexible behavioural learning and that behavioural flexibility itself is unlikely to explain differences in invasion success between these species.

@@

#### Husbandry  
*Breeding colony* -- We tested juveniles originating from a breeding colony established in the lab since 2019. A total of 270 and 180 adults of *L. delicata* and *L. guichenoti*, respectively, were established in containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures had non-stick matting, a shelter, and several small water dishes. Water was given daily, and they were fed approx. 40 mid-size crickets (*Acheta domestica*) per enclosure three days a week. Crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Using a heat chord and a heat lamp we created a temperature gradient and kept lights on a 12 h light:12 h dark cycle. UVA/UVB lighting was also provided. Room temperatures were set to 22-24ºC, and the warm side of enclosures reached 32ºC.

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) with moist vermiculite inside that were placed on one side of the communal enclosures (see above). We checked for the presence of eggs in the boxes three days a week. After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm and weighed them with a digital scale ± 0.001g error (OHAUS, Model spx123). We also assigned identities to the clutch and each egg. Eggs were then treated with CORT or vehicle (see CORT and Temperature manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching.

*Hatchlings* – Eggs in the incubator were checked three times a week for hatchlings. After hatchling, we measured juveniles’ SVL and Tail Length (TL) with a ruler to the nearest mm and weighed them with a digital scale ± 0.001g error. We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with non-stick matting, a shelter, and a small water dish. During this period, they were sprayed with water every day and received 3-6 small _A. domestica_ crickets three times a week. All care otherwise followed similar protocols to adults (see above).
Two weeks before starting the training phase (see below), lizards were moved to an experimental arena for acclimatisation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the side and a water dish on the other. These new enclosures were placed in two rooms across 7 different racks that were monitored by 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). Conditions in these rooms where identical to the main room, but the number of lizards per species and treatment in each rack was counterbalanced to control for any potential effect of the room or the position of the lizard on the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied _ad libitum_. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The temperature of both rooms was set to between 22-24ºC.

@@

To empirically test the effect of early environment we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23ºC ± 3ºC or Hot -- 30ºC ± 3ºC) in a 2x2 factorial design (@fig-Methods A). We first allocated eggs into one of two different treatments: in the CORT treatment, eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (vehicle) at a final concentration of 10 pg CORT/mL; in the Control treatment eggs received an equal volume of the vehicle. This method has been validated before in _L. delicata_ [@crino2024eggs], increasing CORT concentration in eggs by approximately 2 standard deviations above the mean natural concentration. After the hormone treatment, were then incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures [@qualls2000post; @cheetham2011embryonic]. The number of eggs per clutch assigned to each hormone and temperature treatment were counterbalanced in both species.

@@

The habituation phase was divided into three stages: in the first stage, the reward, a small, frozen cricket (*A. domestica*) was placed in an opaque petri dish (3D x 1.6H cm) (@fig-Methods Habituation phase, Stage 1); in the second stage, the petri dish with the cricket was placed on top of the white 3D printed ramps (@fig-Methods Habituation phase, Stage 2); and in the third stage, the cricket was left inside a well (3D x 1.75H cm) on top of the ramp (Fig. @fig-Methods Habituation phase, Stage 3). Trials in every stage lasted one hour, begining when the feeding block (petri dish, ramp, or both) was placed inside the enclosure and finishing when we took the feeding block away. We recorded whether the cricket had been consumed or not and we considered a stage as completed if the lizards ate the crickets in at least 5 out of 6 trials. This phase lasted 38 days, and only in one case a lizard did not progress to the next phases because it did not pass the aqcuisition criteria.

In the colour association (Associative task in Fig. @fig-Methods) and reversal task (Reversal task in @fig-Methods B), we used three ramps that were identical in shape and size but different in the colour. The colours of these ramps were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours [@Baden_Osorio_2019_Vert_vision]. However, to control for potential colour biases, we assigned one group of lizards within each treatment to associate food in the blue ramp followed by the red ramp in the reversal, while in the other group the order was reversed. As in the last stage of the habituation phase, we placed the cricket inside the wells of the ramps, but then two of them were covered with 3D-printed lids (3D x 0.5H cm) so prey was only accessible on one ramp. The food reward was placed on all three ramps to avoid lizards using prey chemical cues that were released in the closed wells through a series of small holes on the top of the lids. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in the association task once a day for 35 days and once a day for 40 days in the reversal.

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using an R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We modelled correct choices [correct (1) or not (0)] as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), along with the three-way interaction between trial, hormone, and temperature. If early environments impact learning then we would predict that the rate of learning (i.e., trial slope) varies by treatment as captured by the interactions. The error structure was modelled using a Bernoulli distribution with a logit link function (family = Bernoulli(link = 'logit')). We included a random intercept and slope (trial) for each lizard in our models. We also incorporated the clutch identity as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple_biology_2014]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design, and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental condition. Learning can also be age-dependent [see @noble_age-dependent_2014], and given that lizards incubated at different temperatures hatch at different times we explored the effects of age on learning. Indeed, lizards’ age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_. However, when this variable was included in the models we did not find any significant effect of age (see Supplementary material). As such, we present models without age as a fixed effect. We used the posterior distributions of parameters from these models to test for differences in learning rate between treatments and species. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates greater than zero were considered as evidence of learning, while those less or equal to zero were not. p~mcmc~ test the hypothesis that slopes and slopes contrasts are different from zero. We considered an effect statistically significant if pmcmc < 0.05.

#### Ethical note
We collected subjects from a breeding colony kept in the lab since 2019. Both the breeding animals and the ones used in this experiment were provided humane laboratory housing, with thermoregulatory opportunities, light (UV and heat) and moderate levels of humidity (see above for details). Lizards in the colony received food ad libitum while the subjects in the experiment were feed one cricket daily during the experiment. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Animals born in captivity cannot be re-released into the wild as stringent biosafety protocols across the states prohibit this. It is also not possible to rehome such a large number of animals. Therefore, upon completion of experimental trials, lizards were kept in their enclosure until they were humanely euthanized following approved procedures. To euthanise animals, we use an injection of sodium pentobarbital (i.e., Lethabarb) (diluted 1:4 ringers solution) at 60-100mg/kg, injected intrapleuroperitoneally. We monitored the animals to ensure there was no irritation from the agent as indicated by distressed animals. Gently pinching the toes was used to confirm the absence of a response before the animal was disposed of.  
Crickets were maintained in big communal enclosures (68.5 L x 49 W x 39 H cm) in the same room where the big colony is, and at the same temperature and light conditions. They were provided vegetables and water ad libitum, and they were given several egg carton shelters to refuge. All lizards were fed alive crickets except during the experiment, when we used crickets frozen at -21 ºC for 24 hours.

@@

We started with 96 lizards, 48 per species and 12 per treatment per species. However, our final sample size was 81 due to natural mortality (n = 11), failure to complete the training stage (n = 1), or because the number of failed tasks (see above) was higher than 15 (n = 3). The final sample sizes per treatment and species are listed in @fig-results.

@@

Both species learnt across all treatments (@fig-results B, D), but we did not find any difference between treatments or species in the rate of learning [mean slopes (denoted as β throughout) per treatment for both species provided in Table 2 in Supplementary Material]. For _L. delicata_, hormone treatment (β~Control~ - β~CORT~ = `r format_dec(mean(c(deli_ControlCold, deli_ControlHot)) - mean(c(deli_CORTCold, deli_CORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(deli_ControlCold, deli_ControlHot) - c(deli_CORTCold, deli_CORTHot)), 3)`), incubation temperature (β~Hot~ - β~Cold~ = `r format_dec(mean(c(deli_CORTHot, deli_ControlHot)) - mean(c(deli_CORTCold, deli_ControlCold)),3)`, p~mcmc~ `r format_p(pmcmc(c(deli_CORTHot, deli_ControlHot) - c(deli_CORTCold, deli_ControlCold)),3)`), or the interaction ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)] = `r format_dec((mean(deli_ControlHot) - mean(deli_CORTHot)) - (mean(deli_ControlCold) - mean(deli_CORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc((deli_ControlHot - deli_CORTHot) - (deli_ControlCold - deli_CORTCold)), 3)`) did not affect learning (see @fig-results A, B). We similarly found no significant effects of  hormone (β~Control~ - β~CORT~ = `r format_dec(mean(c(guich_ControlCold, guich_ControlHot)) - mean(c(guich_CORTCold, guich_CORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(guich_ControlCold, guich_ControlHot) - c(guich_CORTCold, guich_CORTHot)), 3)`), incubation temperature (β~Hot~ - β~Cold~ = `r format_dec(mean(c(guich_CORTHot, guich_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold)),3)`, p~mcmc~ `r format_p(pmcmc(c(guich_CORTHot, guich_ControlHot) - c(guich_CORTCold, guich_ControlCold)),3)`), or the interaction ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)] = `r format_dec((mean(guich_ControlHot) - mean(guich_CORTHot)) - (mean(guich_ControlCold) - mean(guich_CORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc((guich_ControlHot - guich_CORTHot) - (guich_ControlCold - guich_CORTCold)), 3)`) on learning in _L. guichenoti_ (see @fig-results C, D).

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Results for L. delicata (A,B) and L. guichenoti (C, D). Panels A and C show the predicted probability of choosing the correct feeder first over trials. The lines represent the mean predicted probability of choosing the correct feeder first on each trial, and the shaded areas indicate the standard deviation of the mean; both were obtained by using the slope and intercept estimates from the posterior distributions. The different colours indicate the different treatments. Panels B and D show the distribution of the estimates of slopes per each treatment. The x-axis represents the slope estimate, and in the y-axis are the density of the estimates. The different colours indicate the different treatments. Points and bars represent the mean and standard deviation of the mean of the estimates, respectively."
source(here("R", "func.R"))
# Create the dfs for the probability plot A and C) (df_prob)
# L. delicata
df_deliprob <- df_plotAC("deli")
# L. guichenoti
df_guichprob <- df_plotAC("guich")
#
# Create the df for the violin plots in plot B) (df_violin)
# L. delicata
df_deliV <- df_plotBD1("deli")
# L. guichenoti
df_guichV <- df_plotBD1("guich")
#
# Create the df for the points and bars in plot B) (df_points)
# L. delicata
df_deliP <- df_plotBD2("deli")
# L. guichenoti
df_guichP <- df_plotBD2("guich")
#
#
# Make plots deli and guich
source(here("R", "func.R"))
fig_deli <- plotting(sp = "deli", df_prob = df_deliprob, df_violin = df_deliV, df_points = df_deliP)
fig_guich <- plotting(sp = "guich", df_prob = df_guichprob, df_violin = df_guichV, df_points = df_guichP)
# Combine plots of both species into the final figure
fig_results <- plot_grid(fig_deli, fig_guich, ncol = 1)
ggsave("./output/figures/fig_results.png", plot=fig_results, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.png")
```

Finally, when all individuals from all the treatments were pooled together and species were compared, there were not significant differences in the estimated slopes (mean of learning slope in _L. delicata_ = `r format_dec(mean(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot)), 3)`, 95% CI = [`r format_dec(quantile(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot), 0.025), 3)`, `r format_dec(quantile(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot), 0.975), 3)`]; mean learning of slope in _L. guichenoti_ = `r format_dec(mean(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`, 95% CI = [`r format_dec(quantile(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot), 0.025), 3)`, `r format_dec(quantile(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot), 0.975), 3)`]; β~_L. delicata_~ - β~_L. guichenoti_~ = `r format_dec(mean(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot) - c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`). Figures for both species with the raw data were included in the Supplementary Material.

>>>>

## Abstract
Behavioural flexibility, the ability to adjust behaviour adaptively in response to internal or external changes, is expected to be crucial for animals adapting to environmental fluctuations. However, the conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how populations will respond to novel circumstances. Stressful situations faced by the parents can have a direct impact on offspring cognition through the transmission of glucocorticoids - stress-related hormones that affect offspring cognition. At the same time, stressful conditions can influence parental behaviour during nesting and consequently the thermal developmental conditions offspring experience. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. We hypothesized prenatal CORT and cold temperatures would impair performance in the reversal task. Given _L. delicata_’s success as an invasive species, we expected this species to behave more flexibly and be less affected by early environmental conditions. Contrary to our expectations, behavioural flexibility appears to be robust to prenatal temperature and CORT in both species. The lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that novel environments are unlikely to influence flexible behavioural learning and that behavioural flexibility itself is unlikely to explain differences in invasion success between these species.

@@

#### Husbandry  
*Breeding colony* -- We tested juveniles originating from a breeding colony established in the lab since 2019. A total of 270 and 180 adults of _L. delicata_ and _L. guichenoti_, respectively, were established in containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures had non-stick matting, a shelter, and several small water dishes. Water was given daily, and they were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. Crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Using a heat chord and a heat lamp we created a temperature gradient and kept lights on a 12 h light:12 h dark cycle. UVA/UVB lighting was also provided. Room temperatures were set to 22-24ºC, and the warm side of enclosures reached 32ºC.

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) with moist vermiculite inside that were placed on one side of the communal enclosures (see above). We checked for the presence of eggs in the boxes three days a week. After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm and weighed them with a digital scale ± 0.001 g error (OHAUS, Model spx123). We also assigned a unique number to the clutch and each egg. Eggs were then treated with CORT or vehicle (see CORT and Temperature manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching.

*Hatchlings* – Eggs in the incubator were checked three times a week for hatchlings. After hatchling, we measured juveniles’ SVL and Tail Length (TL) with a ruler to the nearest mm and weighed them with a digital scale ± 0.001 g error. We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with non-stick matting, a shelter, and a small water dish. During this period, they were sprayed with water every day and received 3-6 small _A. domestica_ crickets three times a week. All care otherwise followed similar protocols to adults (see above).
Two weeks before starting the training phase (see below), lizards were moved to an experimental arena for acclimatisation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish on the other. These new enclosures were placed in two rooms across 7 different racks that were monitored by 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). Conditions in these rooms where identical to the main room, but the number of lizards per species and treatment in each rack was counterbalanced to control for any potential effect of the room or the position of the lizard on the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied _ad libitum_. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The temperature of both rooms was set to between 22-24 ºC.

@@

To empirically test the effect of early environment we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23 ºC ± 3 ºC or Hot -- 30 ºC ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). We first allocated eggs into one of two different treatments: in the CORT treatment, eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (vehicle) at a final concentration of 10 pg CORT/mL; in the Control treatment eggs received an equal volume of the vehicle. This method has been validated before in _L. delicata_ [@crino2024eggs], increasing CORT concentration in eggs by approximately 2 standard deviations above the mean natural concentration. After the hormone treatment, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures [@qualls2000post; @cheetham2011embryonic].

We counterbalanced the number of eggs assigned to each hormone and temperature treatment using a partial split-clutch design. We assigned at least one egg per clutch to each treatment, and the remaining eggs were randomly assigned to one of the treatments. When the eggs per clutch was less than four eggs, we assigned each egg randomly to one of the treatments. The final total number of clutches per species is included in Results.

@@

The habituation phase was divided into three stages: in the first stage, the reward, a small, frozen cricket (*A. domestica*) was placed in an opaque petri dish (3 D x 1.6 H cm) (@fig-Methods Habituation phase, Stage 1); in the second stage, the petri dish with the cricket was placed on top of the white 3D printed ramps (@fig-Methods Habituation phase, Stage 2); and in the third stage, the cricket was left inside a well (3D x 1.75H cm) on top of the ramp (Fig. @fig-Methods Habituation phase, Stage 3). Trials in every stage lasted one hour, beginning when the feeding block (petri dish, ramp, or both) was placed inside the enclosure and finishing when we took the feeding block away. We recorded whether the cricket had been consumed or not and we considered a stage as completed if the lizards ate the crickets in at least 5 out of 6 trials. This phase lasted 38 days, and only in one case a lizard did not progress to the next phases because it did not pass the aqcuisition criteria.

In the colour association (Associative task in @fig-Methods) and reversal task (Reversal task in @fig-Methods B), we used three ramps that were identical in shape and size but different in the colour. The colours of these ramps were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours [@Baden_Osorio_2019_Vert_vision]. However, to control for potential colour biases, we assigned one group of lizards within each treatment to associate food in the blue ramp followed by the red ramp in the reversal, while in the other group the order was reversed. As in the last stage of the habituation phase, we placed the cricket inside the wells of the ramps, but then two of them were covered with 3D-printed lids (3 D x 0.5 H cm) so prey was only accessible on one ramp. The food reward was placed on all three ramps to avoid lizards using prey chemical cues that were released in the closed wells through a series of small holes on the top of the lids. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in the association task once a day for 35 days and once a day for 40 days in the reversal.

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using an R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We modelled correct choices [correct (1) or not (0)] as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), along with the three-way interaction between trial, hormone, and temperature. If early environments impact learning then we would predict that the rate of learning (i.e., trial slope) varies by treatment as captured by the interactions. The error structure was modelled using a Bernoulli distribution with a logit link function (family = Bernoulli(link = 'logit')). We included a random intercept and slope (trial) for each lizard in our models. We also incorporated the clutch identity as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple_biology_2014]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design, and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

Learning can also be age-dependent [see @noble_age-dependent_2014], and given that lizards incubated at different temperatures hatch at different times we explored the effects of age on learning. Indeed, lizards’ age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_. However, when this variable was included in the models we did not find any significant effect of age (see Supplementary Material). As such, we present models without age as a fixed effect.

We used the posterior distributions of parameters from these models to test for differences in learning rate between treatments and species. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates greater than zero were considered as evidence of learning, while those less or equal to zero were not. p~mcmc~ test the hypothesis that slopes and slopes contrasts are different from zero. We considered an effect statistically significant if p~mcmc~ < 0.05.

#### Ethical note
We collected subjects from a breeding colony kept in the lab since 2019. Both the breeding animals and the ones used in this experiment were provided humane laboratory housing, with thermoregulatory opportunities, light (UV and heat) and moderate levels of humidity (see above for details). Lizards in the colony received food _ad libitum_ while the subjects in the experiment were fed one cricket daily during the experiment. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Animals born in captivity cannot be re-released into the wild as stringent biosafety protocols across the states prohibit this. It is also not possible to rehome such a large number of animals. Therefore, upon completion of experimental trials, lizards were kept in their enclosure until they were humanely euthanized following approved procedures. To euthanise animals, we use an injection of sodium pentobarbital (i.e., Lethabarb) (diluted 1:4 ringers solution) at 60-100mg/kg, injected intrapleuroperitoneally. We monitored the animals to ensure there was no irritation from the agent as indicated by distressed animals. Gently pinching the toes was used to confirm the absence of a response before the animal was disposed of.  
Crickets were maintained in big communal enclosures (68.5 L x 49 W x 39 H cm) in the same room where the big colony is, and at the same temperature and light conditions. They were provided vegetables and water ad libitum, and they were given several egg carton shelters to refuge. All lizards were fed alive crickets except during the experiment, when we used crickets frozen at -21 ºC for 24 hours.

@@

```{r, countclutches}
#| label: countclutches
# Count the number of clutches per species
## L. delicata
deli_clutches <- clean_df %>% filter(species == "delicata") %>% distinct(clutch) %>% nrow()
## L. guichenoti
guich_clutches <- clean_df %>% filter(species == "guichenoti") %>% distinct(clutch) %>% nrow()
```

We started with 96 lizards, 48 per species and 12 per treatment per species. However, our final sample size was 81 due to natural mortality (n = 11), failure to complete the training stage (n = 1), or because the number of failed tasks (see above) was higher than 15 (n = 3). The final sample sizes per treatment and species are listed in @fig-results. These animals came from a total of `r deli_clutches` clutches in _L. delicata_ and `r guich_clutches` in _L. guichenoti_.

@@

Both species learnt across all treatments (@fig-results B, D), but we did not find any difference between treatments or species in the rate of learning [mean slopes (denoted as β throughout) per treatment for both species provided in Table 1 in Supplementary Material]. For _L. delicata_, hormone treatment (β~Control~ - β~CORT~ = `r format_dec(mean(c(deli_ControlCold, deli_ControlHot)) - mean(c(deli_CORTCold, deli_CORTHot)), 3)`, 95% CI = [`r format_dec(quantile(c(deli_ControlCold, deli_ControlHot) - c(deli_CORTCold, deli_CORTHot), 0.025), 3)`, `r format_dec(quantile(c(deli_ControlCold, deli_ControlHot) - c(deli_CORTCold, deli_CORTHot), 0.975), 3)`], p~mcmc~ `r format_p(pmcmc(c(deli_ControlCold, deli_ControlHot) - c(deli_CORTCold, deli_CORTHot)), 3)`), incubation temperature (β~Hot~ - β~Cold~ = `r format_dec(mean(c(deli_CORTHot, deli_ControlHot)) - mean(c(deli_CORTCold, deli_ControlCold)),3)`, 95% CI = [`r format_dec(quantile(c(deli_CORTHot, deli_ControlHot) - c(deli_CORTCold, deli_ControlCold), 0.025), 3)`, `r format_dec(quantile(c(deli_CORTHot, deli_ControlHot) - c(deli_CORTCold, deli_ControlCold), 0.975), 3)`], p~mcmc~ `r format_p(pmcmc(c(deli_CORTHot, deli_ControlHot) - c(deli_CORTCold, deli_ControlCold)),3)`), or the interaction ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)] = `r format_dec((mean(deli_ControlHot) - mean(deli_CORTHot)) - (mean(deli_ControlCold) - mean(deli_CORTCold)), 3)`, 95%CI = [`r format_dec(quantile((deli_ControlHot - deli_CORTHot) - (deli_ControlCold - deli_CORTCold), 0.025), 3)`, `r format_dec(quantile((deli_ControlHot - deli_CORTHot) - (deli_ControlCold - deli_CORTCold), 0.975), 3)`], p~mcmc~ `r format_p(pmcmc((deli_ControlHot - deli_CORTHot) - (deli_ControlCold - deli_CORTCold)), 3)`) did not affect learning (see @fig-results A, B). We similarly found no significant effects of  hormone (β~Control~ - β~CORT~ = `r format_dec(mean(c(guich_ControlCold, guich_ControlHot)) - mean(c(guich_CORTCold, guich_CORTHot)), 3)`, 95% CI = [`r format_dec(quantile(c(guich_ControlCold, guich_ControlHot) - c(guich_CORTCold, guich_CORTHot), 0.025), 3)`, `r format_dec(quantile(c(guich_ControlCold, guich_ControlHot) - c(guich_CORTCold, guich_CORTHot), 0.975), 3)`], p~mcmc~ `r format_p(pmcmc(c(guich_ControlCold, guich_ControlHot) - c(guich_CORTCold, guich_CORTHot)), 3)`), incubation temperature (β~Hot~ - β~Cold~ = `r format_dec(mean(c(guich_CORTHot, guich_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold)),3)`, 95% CI = [`r format_dec(quantile(c(guich_CORTHot, guich_ControlHot) - c(guich_CORTCold, guich_ControlCold), 0.025), 3)`, `r format_dec(quantile(c(guich_CORTHot, guich_ControlHot) - c(guich_CORTCold, guich_ControlCold), 0.975), 3)`], p~mcmc~ `r format_p(pmcmc(c(guich_CORTHot, guich_ControlHot) - c(guich_CORTCold, guich_ControlCold)),3)`), or the interaction ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)] = `r format_dec((mean(guich_ControlHot) - mean(guich_CORTHot)) - (mean(guich_ControlCold) - mean(guich_CORTCold)), 3)`, 95%CI = [`r format_dec(quantile((guich_ControlHot - guich_CORTHot) - (guich_ControlCold - guich_CORTCold), 0.025), 3)`, `r format_dec(quantile((guich_ControlHot - guich_CORTHot) - (guich_ControlCold - guich_CORTCold), 0.975), 3)`], p~mcmc~ `r format_p(pmcmc((guich_ControlHot - guich_CORTHot) - (guich_ControlCold - guich_CORTCold)), 3)`) on learning in _L. guichenoti_ (see @fig-results C, D).

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Results for L. delicata [(a), (b)] and L. guichenoti [(c), (d)]. Panels (a) and (c) show the predicted probability of choosing the correct feeder first over trials. The lines represent the mean predicted probability of choosing the correct feeder first on each trial, and the shaded areas indicate the standard deviation of the mean; both were obtained by using the slope and intercept estimates from the posterior distributions. The different colours indicate the different treatments. Panels (b) and (d) show the distribution of the estimates of slopes per each treatment. The x-axis represents the slope estimate, and in the y-axis are the density of the estimates. The different colours indicate the different treatments. Points and bars represent the mean and standard deviation of the mean of the estimates, respectively."
source(here("R", "func.R"))
# Create the dfs for the probability plot A and C) (df_prob)
# L. delicata
df_deliprob <- df_plotAC("deli")
# L. guichenoti
df_guichprob <- df_plotAC("guich")
#
# Create the df for the violin plots in plot B) (df_violin)
# L. delicata
df_deliV <- df_plotBD1("deli")
# L. guichenoti
df_guichV <- df_plotBD1("guich")
#
# Create the df for the points and bars in plot B) (df_points)
# L. delicata
df_deliP <- df_plotBD2("deli")
# L. guichenoti
df_guichP <- df_plotBD2("guich")
#
#
# Make plots deli and guich
source(here("R", "func.R"))
fig_deli <- plotting(sp = "deli", df_prob = df_deliprob, df_violin = df_deliV, df_points = df_deliP)
fig_guich <- plotting(sp = "guich", df_prob = df_guichprob, df_violin = df_guichV, df_points = df_guichP)
# Combine plots of both species into the final figure
fig_results <- plot_grid(fig_deli, fig_guich, ncol = 1)
ggsave("./output/figures/fig_results.png", plot=fig_results, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.png")
```
